---
output:
  rmarkdown::github_document
bibliography: "inst/REFERENCES.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R/`scPCA`

[![Travis-CI Build Status](https://travis-ci.org/PhilBoileau/scPCA.svg?branch=master)](https://travis-ci.org/PhilBoileau/scPCA)
[![AppVeyor Build  Status](https://ci.appveyor.com/api/projects/status/github/PhilBoileau/scPCA?branch=master&svg=true)](https://ci.appveyor.com/project/PhilBoileau/scPCA/)
[![Coverage Status](https://img.shields.io/codecov/c/github/PhilBoileau/scPCA/master.svg)](https://codecov.io/github/PhilBoileau/scPCA?branch=master)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)
<!--
[![BioC status](http://www.bioconductor.org/shields/build/release/bioc/scPCA.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/scPCA)
[![Bioc Time](http://bioconductor.org/shields/years-in-bioc/scPCA.svg)](https://bioconductor.org/packages/release/bioc/html/scPCA.html)
[![Bioc Downloads](http://bioconductor.org/shields/downloads/scPCA.svg)](https://bioconductor.org/packages/release/bioc/html/scPCA.html)
-->

> Sparse Contrastive Principal Component Analysis for Computational Biology

__Authors:__ [Philippe Boileau](https://pboileau.ca/),
  [Nima Hejazi](https://nimahejazi.org), William Krinsman

---

## What's `scPCA`?

The exploration and analysis of modern high-dimensional biological data
regularly involves the use of dimension reduction techniques in order to tease
out meaningful and interpretable information from complex experimental data,
often subject to batch effects and other technical noise. In tandem with the
development of sequencing technology (e.g., RNA-seq, scRNA-seq), many variants
of PCA have been developed in attempts to remedy deficiencies in
interpretability and stability that plague their classical variants. Such
developments have included both sparse PCA (SPCA) [@zou2006sparse], which
increases the stability and interpretability of principal component loadings in
high dimensions, and, more recently, contrastive PCA (cPCA)
[@abid2018exploring], which captures relevant information in the target
(experimental) data set by eliminating technical noise through comparison to a
so-called background data set. While SPCA and cPCA have both individually proven
useful in resolving distinct shortcomings of PCA, neither is capable of
simultaneously tackling the issues of interpretability, stability and relevance
simultaneously. The `scPCA` package implements _sparse constrastive PCA_, to
accomplish both of these tasks in the context of high-dimensional biological
data. In addition to implementing this newly developed technique, the `scPCA`
package also implements cPCA and generalizations of it.

<!--
...
-->

---

## Installation

<!-- For standard use, install from -->
<!-- [Bioconductor](https://bioconductor.org/packages/scPCA) using -->
<!-- [`BiocManager`](https://CRAN.R-project.org/package=BiocManager): -->

<!-- ```{r bioc-installation, eval = FALSE} -->
<!-- if (!requireNamespace("BiocManager", quietly=TRUE)) { -->
<!--   install.packages("BiocManager") -->
<!-- } -->
<!-- BiocManager::install("scPCA") -->
<!-- ``` -->

To contribute, install the bleeding-edge _development version_ from GitHub via
[`remotes`](https://CRAN.R-project.org/package=remotes):

```{r gh-master-installation, eval = FALSE}
remotes::install_github("PhilBoileau/scPCA")
```

<!-- Current and prior [Bioconductor](https://bioconductor.org) releases are -->
<!-- available under branches with numbers prefixed by "RELEASE_". For example, to -->
<!-- install the version of this package available via Bioconductor 3.10, use -->

<!-- ```{r gh-develop-installation, eval = FALSE} -->
<!-- devtools::install_github("PhilBoileau/scPCA", ref = "RELEASE_3_10") -->
<!-- ``` -->

---

## Example

For details on how to best use the `scPCA` R package, please consult the most
recent vignette.
<!-- [package -->
<!-- vignette](https://bioconductor.org/packages/release/bioc/vignettes/scPCA/inst/doc/scpca_intro.html) -->
<!-- available through the [Bioconductor -->
<!-- project](https://bioconductor.org/packages/scPCA). -->

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/PhilBoileau/scPCA/issues).

---

## Contributions

Contributions are very welcome. Interested contributors should consult our
[contribution
guidelines](https://github.com/PhilBoileau/scPCA/blob/master/CONTRIBUTING.md)
prior to submitting a pull request.

---

## License

&copy; 2019 [Philippe Boileau](https://pboileau.ca/)

The contents of this repository are distributed under the MIT license. See file
`LICENSE` for details.

---

## References

