---
output:
  rmarkdown::github_document
bibliography: "inst/REFERENCES.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R/`scPCA`

[![Travis-CI Build Status](https://travis-ci.org/PhilBoileau/scPCA.svg?branch=master)](https://travis-ci.org/PhilBoileau/scPCA)
[![AppVeyor Build  Status](https://ci.appveyor.com/api/projects/status/github/PhilBoileau/scPCA?branch=master&svg=true)](https://ci.appveyor.com/project/PhilBoileau/scPCA/)
[![Coverage Status](https://img.shields.io/codecov/c/github/PhilBoileau/scPCA/master.svg)](https://codecov.io/github/PhilBoileau/scPCA?branch=master)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![BioC status](http://www.bioconductor.org/shields/build/release/bioc/scPCA.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/scPCA)
[![Bioc Time](http://bioconductor.org/shields/years-in-bioc/scPCA.svg)](https://bioconductor.org/packages/release/bioc/html/scPCA.html)
[![Bioc Downloads](http://bioconductor.org/shields/downloads/scPCA.svg)](https://bioconductor.org/packages/release/bioc/html/scPCA.html)
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](http://opensource.org/licenses/MIT)

> Sparse and Contrastive Principal Components Analysis for Computational Biology

__Authors:__ [Philippe Boileau](https://github.com/PhilBoileau),
  [Nima Hejazi](https://nimahejazi.org), William Krinsman

---

## What's `scPCA`?

The `scPCA` R package facilitates
@abid2018exploring

---

## Installation

For standard use, install from
[Bioconductor](https://bioconductor.org/packages/scPCA) using
[`BiocManager`](https://CRAN.R-project.org/package=BiocManager):

```{r bioc-installation, eval = FALSE}
if (!requireNamespace("BiocManager", quietly=TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("scPCA")
```

To contribute, install the bleeding-edge _development version_ from GitHub via
[`devtools`](https://www.rstudio.com/products/rpackages/devtools/):

```{r gh-master-installation, eval = FALSE}
devtools::install_github("PhilBoileau/scPCA")
```

Current and prior [Bioconductor](https://bioconductor.org) releases are
available under branches with numbers prefixed by "RELEASE_". For example, to
install the version of this package available via Bioconductor 3.10, use

```{r gh-develop-installation, eval = FALSE}
devtools::install_github("PhilBoileau/scPCA", ref = "RELEASE_3_10")
```

---

## Example

For details on how to best use the `scPCA` R package, please consult the most
recent [package
vignette](https://bioconductor.org/packages/release/bioc/vignettes/scPCA/inst/doc/scpca_intro.html)
available through the [Bioconductor
project](https://bioconductor.org/packages/scPCA).

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/PhilBoileau/scPCA/issues).

---

## Contributions

Contributions are very welcome. Interested contributors should consult our
[contribution
guidelines](https://github.com/PhilBoileau/scPCA/blob/master/CONTRIBUTING.md)
prior to submitting a pull request.

---

## License

&copy; 2019 [Philippe Boileau](https://github.com/PhilBoileau)

The contents of this repository are distributed under the MIT license. See file
`LICENSE` for details.

---

## References

